---
title: "Routine maintenance and CI"
description: "Update schemas, run thin maintenance commands, inspect reports, and publish client docs."
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Routine maintenance and CI}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

Use the same loop for new and existing clients: **update inputs -> plan -> review
-> apply -> check -> test -> commit**. Schemas, service policy, callbacks, fixed
fixtures, output, and `.specmill/manifest.json` form one reviewed change.

## 1. Review schema changes

Keep the old schema snapshot while downloading the new one using your client's
own acquisition process. Use the same filename and selection policy for both
versions. For example, place `catalogue.json` in `schema-before/` and `schema-after/`:

```{r compare, eval=FALSE}
selection <- list(methods = c('GET', 'POST'), exclude = '^/internal/')
diff <- specmill::schema_diff(
  '/path/to/schema-before', '/path/to/schema-after',
  policies = list('catalogue.json' = selection)
)
cat(specmill::format_diff_markdown(diff))
specmill::count_diff_changes(diff)
```

`schema_diff()` compares method/path identities and canonical data including
reachable local references and response shapes. Changed contracts enter the
breaking-change report lane conservatively for review. Parsing failures stop
comparison. These labels do not prove runtime compatibility.

For parsed operation records, `compare_operations(old, new)` gives a smaller
report of additions, removals, changed parameters/bodies, and unknown response
compatibility. Unsupported operations remain parser diagnostics, so review them
alongside any comparison.

Compare schema revisions under a fixed policy, then review policy changes
separately. Otherwise, changing an exclusion can look like an upstream deletion.
Generation does not automatically delete an implementation merely because an
operation disappears from an unsupported or changed schema. Eligible removal
comes from explicit exclusions or reconciled renames and requires verified ownership.

## 2. Generate and verify

```{r loop, eval=FALSE}
root <- normalizePath('/path/to/client', winslash = '/', mustWork = TRUE)
plan <- specmill::generate_client(root, config = 'specmill.yml', mode = 'plan')
plan$files
plan$drift
plan$diagnostics
# After reviewing the change:
specmill::generate_client(root, config = 'specmill.yml', mode = 'apply')
specmill::generate_client(root, config = 'specmill.yml', mode = 'check')
testthat::test_local(root, filter = 'contract-', stop_on_failure = TRUE)
```

Supply the same explicit `callbacks` environment on each call when your policy
uses callbacks. Run the relevant manual/helper tests too. Reapplying unchanged
inputs should produce only unchanged/retained file actions. Review output diffs
before committing; do not hand-edit generated files to fix a policy error.

## 3. Add a thin sourceable command

Save the following as `dev/generate.R` in a client without an existing command:

```r
.client_root <- specmill::script_root('generate.R')
generate_main <- function(root = .client_root, args = commandArgs(trailingOnly = TRUE)) {
  specmill::generation_command(root, args = args, kind = 'stubs')
}
if (sys.nframe() == 0L) generate_main()
```

`script_root()` locates the package from the source frame or Rscript invocation,
so the command works from another working directory. Add explicit callback
loading inside `generate_main()` if needed, as in the existing-client guide.
Source it in R and call `generate_main(args = '--plan')`, or run:

```sh
Rscript dev/generate.R --plan
Rscript dev/generate.R --check
Rscript dev/generate.R
```

The final command **applies**. Compatibility commands intentionally have a
different default from the core `generate_client()`, which defaults to check.

| Command kind | Preview | Check | Apply |
|---|---|---|---|
| `stubs` (wrappers and documentation) | `--plan` | `--check` | No mode flag |
| `tests` (fixed contract tests) | `--dry-run` | `--check` | `--generate` or no mode flag |

`--help` describes accepted flags. The stubs command accepts legacy
`--rebuild=<prefix>` flags; YAML still controls selection. The tests command
accepts `--force`, which never overrides ownership. Check/apply test commands
fail if any selected operation lacks a fixed contract. Plan can report the gaps.

For custom orchestration, use `generate_client(artifacts = ...)` to limit
reconciliation to `wrappers`, `tests`, or `documentation`. Shared ownership and
manifest checks still apply.

## 4. Inspect implementation and test gaps

```{r inspect, eval=FALSE}
inspection <- specmill::inspect_client(root)
inspection$coverage
inspection$manual_exports
inspection$diagnostics
```

Coverage counts files and declarations, not passing tests or working endpoints.
Selected unsupported operations remain in totals. Manual exported wrappers are
separate from the schema denominator and still need their own tests.

For badges and baseline comparisons, create a separate client report policy:

```yaml
# dev/coverage.yml
baseline: dev/coverage-baseline.json
groups:
  catalogue:
    services: '^catalogue$'
    label: Catalogue wrappers
    badge: man/figures/catalogue-coverage.json
```

Group `services` values are regex strings matching service IDs. Groups must
partition every configured service exactly once. Create the baseline's parent
directory before apply. Report-policy filenames passed as strings are resolved
by R against the working directory; make them explicit with `file.path(root, ...)`:

```{r reports, eval=FALSE}
specmill::coverage_report(root, file.path(root, 'dev/coverage.yml'))
specmill::coverage_report(root, file.path(root, 'dev/coverage.yml'), mode = 'apply')
specmill::test_gap_report(root, list(
  helpers = list('api_request'), report_dir = 'dev/test-gaps',
  readiness_report = 'dev/readiness-report.json'
))
```

Reports default to read-only plans. Coverage apply writes badges and its baseline;
gap-report apply writes a dated JSON report. Gap reasons distinguish missing
contracts, absent/empty test files, and unsupported operations. `readiness_report`
is a pointer to your canonical audit, not an audit created by this call.

## 5. Add client-specific checks to CI

CI should install the project's reviewed toolkit and formatter, then run
generation checks before tests. For a client with no callbacks, an R CI step can be:

```r
checked <- specmill::generate_client('.', config = 'specmill.yml', mode = 'check')
stopifnot(length(checked$diagnostics) == 0L)
testthat::test_local('.', stop_on_failure = TRUE)
```

If you intentionally retain unsupported operations, use an explicit reviewed
diagnostic allowance instead of removing the diagnostics gate. Do not use
`continue-on-error` for invalid configuration or schema parsing before publishing
generated output. Generation commands write existing summary keys to
`GITHUB_OUTPUT` when that environment variable is set; core functions return
structured results and errors instead of choosing a process exit code.

`check_public_boundary()` accepts your forbidden host/file/export/schema regexes
and approved manual-export list. `check_client_hooks()` validates supplied runtime
hook declarations against parsed client definitions. Their reference pages explain
arguments and return values. Keep these policies client-specific.

`credential_status()` and `credential_preflight()` detect absent/placeholder
credentials without logging values or making requests. They cannot prove a key
is authorized. Offline generation and fixed contract tests need no live key.

## 6. Build documentation for your generated client

Set `documentation: true` and write useful `docs` policy before building the
client site. Generation stages roxygen and reconciles `man/` and `NAMESPACE`;
the help pages then become pkgdown reference pages. Add human-written vignettes
for authentication, common workflows, result interpretation, and errors.

From the **client package root**, install the documentation tools and initialize
pkgdown once:

```{r client-site, eval=FALSE}
install.packages(c('pkgdown', 'usethis', 'rmarkdown', 'knitr'))
usethis::use_pkgdown()
pkgdown::build_site(preview = FALSE)
```

For a GitHub-hosted client, `usethis::use_pkgdown_github_pages()` configures the
publishing workflow and Pages settings. Review the generated workflow for your
default branch. See [usethis's pkgdown setup](https://usethis.r-lib.org/reference/use_pkgdown.html).

Keep its README short: purpose, installation, one example, and links to guides.
Put sequential workflows in vignettes and per-function detail in reference docs.
Render examples offline; mark real API calls `eval=FALSE` or use fixed examples
that do not contact services.

For **specmill's own site**, `_pkgdown.yml` defines navigation and
`.github/workflows/pkgdown.yml` builds on pull requests and publishes on the
configured default-branch pushes or manual dispatch. Local build command:

```{r specmill-site, eval=FALSE}
pkgdown::build_site(preview = FALSE)
```

Open `docs/index.html` after building. Site output is ignored by Git and package
builds; source articles and help pages are tracked. The workflow follows the
[r-lib pkgdown example](https://github.com/r-lib/actions/blob/v2/examples/pkgdown.yaml).
