---
title: "Write and run contract tests"
description: "Create independent request fixtures, generate tests, and understand what they prove."
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Write and run contract tests}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = '#>', message = FALSE)
```

A generated wrapper can call the right helper and still return the wrong result
or throw afterward. Fixed contract tests check successful completion, exact
helper arguments, call order/count, and the final result. Write expectations
from the reviewed API/helper contract, independently of generated wrapper text.

This standalone example creates a temporary client and runs its generated test
offline. Install `httr2` and `testthat` first if they are not available.

## 1. Prepare a client with one selected operation

```{r client}
schema <- system.file('catalogue/schema.json', package = 'specmill', mustWork = TRUE)
root <- tempfile('tested-client-')
created <- specmill::initialize_client(
  root, schema, package = 'testedclient', title = 'Tested Catalogue Client',
  author = list(given = 'Example', family = 'Maintainer', email = 'you@example.org'),
  license = 'MIT + file LICENSE', base_url = 'https://api.example.org'
)
cat('\nselection:\n  methods: [GET]\n  exclude: ["^/items$"]\n',
    file = file.path(root, 'apis/default.yml'), append = TRUE)
```

For your real client, use its existing root and selection policy. Do not narrow
selection simply to hide missing tests. Here the teaching client intentionally
has just `get_item()`.

## 2. Save independent fixed expectations

An RDS contract preserves R types and classes. `calls` is an ordered sequence;
each entry contains a helper name, its expected arguments, and its mock response.

```{r fixture}
contracts <- list(
  get_item = list(
    inputs = list(item_id = 'item-1', language = 'en'),
    calls = list(list(
      helper = 'api_request',
      arguments = list(
        method = 'GET', path = '/items/{item_id}',
        path_params = list(item_id = 'item-1'),
        query = list(language = 'en'), body = NULL
      ),
      response = list(id = 'item-1', title = 'Example item')
    )),
    result = list(id = 'item-1', title = 'Example item')
  )
)
fixture <- file.path(root, 'tests/testthat/fixtures/contracts.rds')
dir.create(dirname(fixture), recursive = TRUE)
saveRDS(contracts, fixture)
cat('\ncontracts_file: tests/testthat/fixtures/contracts.rds\n',
    file = file.path(root, 'apis/default.yml'), append = TRUE)
```

Keys such as `get_item` are public wrapper names. Inputs use their public R names.
Specify types intentionally: `1L` is an integer, `1` is a double, and exact
expectations distinguish them. Use actual response classes (for example tibbles)
when preserving an existing client. Never put credentials in fixtures.

Optional contract `environment` values are scoped environment variables for the
test, implemented with `withr::local_envvar()`. Add `withr` to that client's test
dependencies if using this field. Keep environment values nonsecret.

## 3. Add the standard package test runner

Initialization declares testthat in `Suggests` but does not create its runner.
For this new client, create `tests/testthat.R`:

```{r runner}
writeLines(c(
  'library(testthat)', 'library(testedclient)', "test_check('testedclient')"
), file.path(root, 'tests/testthat.R'))
```

Use your own package name. Existing packages should keep their current runner
and manual tests.

## 4. Generate and run the test

```{r generate}
plan <- specmill::generate_client(root, config = 'specmill.yml', mode = 'plan')
stopifnot(!length(plan$diagnostics), length(plan$operations) == 1L)
applied <- specmill::generate_client(root, config = 'specmill.yml', mode = 'apply')
checked <- specmill::generate_client(root, config = 'specmill.yml', mode = 'check')
stopifnot(file.exists(file.path(root, 'tests/testthat/test-contract-get_item.R')))
```

The fixed sequence test reads the RDS fixture and replaces `api_request` in the
client namespace with a recorder. The placeholder URL is never contacted.

```{r run-tests}
testthat::test_local(root, filter = 'contract-get_item', stop_on_failure = TRUE)
```

Use `stop_on_failure = TRUE` in automation. Merely exiting R successfully is not
enough if a test command only prints failures. In a larger client, also run the
existing bespoke tests for the helpers and hooks you changed.

## 5. Inspect test coverage separately from execution

```{r coverage}
inspection <- specmill::inspect_client(root)
inspection$coverage
stopifnot(inspection$coverage$testedclient$contracts == 1L)
```

`implemented` counts selected named exports with source definitions. `contracts`
counts fixed contract declarations with generated test files. Neither count says
tests passed. Unsupported selected operations stay in the coverage denominator;
manual exports are reported separately.

`generation_command(root, kind = 'tests', args = '--check')` also requires fixed
contracts for all selected operations. Its `--dry-run` mode can show gaps before
you finish authoring them. The core `generate_client()` generates supplied
contracts without requiring every selected operation to have one.

## Inline YAML contracts

For a simple single-call test, use `contracts` instead of an RDS file:

```yaml
response_fixture: {id: item-1}
contracts:
  get_item:
    inputs: {item_id: item-1, language: en}
    request:
      method: GET
      path: /items/{item_id}
      path_params: {item_id: item-1}
      query: {language: en}
      body: null
    result: {id: item-1}
```

A contract can override `response_fixture`. Inline single-call tests use
`test-<operation>.R`; RDS sequence tests use `test-contract-<operation>.R`.
Prefer the latter when a manual test already uses the simple name, multiple
helper calls matter, or results have R classes. Do not declare the same operation
in both inline and RDS contracts.

## Candidate input fixtures are not independent expectations

`operation_fixtures()` chooses candidate inputs in the order **override ->
example -> default -> enum -> type fixture**. It checks supported constraints
and stops if the chosen candidate fails them; it does not silently try a lower
priority value. Schema examples are never evaluated as R code.

```{r candidates}
operations <- specmill::read_operations(schema)$operations
candidate_inputs <- specmill::operation_fixtures(operations)
candidate_inputs$get_item
```

Review these values before using them against a service. They do not establish
expected helper calls or correct upstream results. `check_requests()` is a small
alternative for checking one zero-argument call against a capture function and
independent result; see its reference example.

## Add the verification layer your change needs

| Layer | What to verify |
|---|---|
| Generation check | Committed output matches schema, policy, callbacks, and fixtures |
| Fixed helper contract | Successful wrapper result plus exact ordered helper calls |
| Hook tests | Validation, parameter changes, skip behavior, and post-processing |
| Local transport test | Final method, encoded URL, query/body placement, headers, media, and error decoding |
| Bounded live check | Current endpoint/authentication behavior using reviewed nonsecret input |

Use a local HTTP server or existing HTTP fixtures to test transport. Do not
re-record production traffic just to update generated tests. A live failure can
come from input, authentication, schema drift, generation, runtime code, or an
outage. Diagnose it before editing generated source or relaxing expectations.
