---
title: "Supported schemas, file safety, and troubleshooting"
description: "Interpret diagnostics, resolve protected output, and recover interrupted application."
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Supported schemas, file safety, and troubleshooting}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Supported input and transport

apipak reads local JSON documents in OpenAPI 3.0/3.1 or Swagger 2.0 format.
YAML project configuration does not imply support for YAML schema documents.
The parser is a supported-subset extractor, not a full OpenAPI validator.

| Feature | Current behavior |
|---|---|
| Scalar path/query parameters | Supported; path parameters must be required |
| Local `$ref` | Supported within a document; external or cyclic input references are diagnosed |
| JSON scalar bodies | Supported |
| Nested JSON objects/arrays | Supported with declared properties and supported item types |
| Array/object path/query inputs | Outside the default serializer subset; diagnose or use a complete client mapping |
| Composed/free-form input schemas | Diagnosed when unsupported |
| Multipart, uploads, non-JSON request media | Outside the generated default transport subset |
| HTTP method | Explicitly declared; no inferred GET/POST conversion |
| Pagination/batching | Client behavior; a `page` argument alone causes one request |
| Response interpretation | Client helper owns decoding and service-specific meaning |

Fixture selection checks supported type, enum, numeric bounds, string lengths,
and patterns. Helpers still own complete runtime validation and serialization.
Do not interpret a chosen fixture as a valid live identifier or a safe operation
to invoke. A GET endpoint may itself submit work.

`read_operations()` returns supported `operations`, structurally valid but
unsupported `unsupported_operations`, an `inventory`, and `diagnostics`.
A complete public `inputs` plus `request` mapping can handle supported metadata
through a client serializer. Generation labels that operation `client-mapped`
and retains the original reasons in `mapping_diagnostics`. Malformed metadata
or broken references do not become safe merely by adding a mapping.

`implementation: existing` retains a declared manual implementation and reports
limitations in `retained_diagnostics`. Unsupported selections stay visible in
coverage totals. They never authorize deleting old output.

## Common errors

| Symptom | Next step |
|---|---|
| Package/function not found | Install the reviewed toolkit, restart R, and inspect `packageVersion('apipak')` and `find.package('apipak')` |
| Missing schema / pattern matched nothing | Resolve paths against `root`, check case, and use sequences such as `files: [schema.json]` |
| Unknown YAML field / invalid regex | Compare with [Configuration](configuration.html); use case-sensitive stringr syntax |
| Unknown operation override | Match the original method/path, including path parameter spelling and trailing slash |
| Name collision | Supply distinct public names across all services, including case; avoid helper/callback names |
| Unresolved callback | Source your development callbacks into the explicitly supplied environment |
| Missing helper definition / helper argument | Define it in client `R/` and map the actual helper signature |
| Missing fixed contracts | Author independent fixtures for selected operations; use test `--dry-run` to inspect gaps |
| Generated output is stale or protected | Inspect a fresh plan; apply reviewed input changes or resolve ownership |
| Formatter version mismatch | Install the exact Air version declared by the project; keep formatting policy consistent |
| Unsafe documentation tag | Put executable/custom roxygen handlers in client code, and use `docs` fields for policy |
| Plan/input changed during generation | Stop concurrent edits/writers, make callbacks deterministic, then replan |
| HTTP failure | Diagnose authentication, input, request construction, schema drift, runtime behavior, and service status separately |

A normal return from `check` only establishes freshness for reconciled output.
Inspect diagnostics and run tests separately. For reproducible generation, pin
the toolkit and formatter alongside schema, policy, and fixture versions.

## Ownership and protected files

`.apipak/manifest.json` records output hashes and generation input metadata.
Keep it in Git. A header alone cannot grant ownership. apipak protects unowned
files, locally edited output, and protected lifecycle states. It rejects groups
containing undeclared definitions or other top-level code.

Use [reviewed adoption](existing-clients.html) only after comparing a legacy file
with the intended generated contract. Never compute adoption hashes across every
file merely to silence protection. If your local change is intentional, move it
into policy or a maintained callback, or retain that wrapper explicitly.

For a rename, the protected original blocks replacement so the old public
function cannot silently remain beside a new one. Explicit exclusions can remove
verified owned output; protected removals are reported as retained. Review the
file actions rather than assuming all requested deletions happened.

## Interrupted apply and recovery

Application parses R output and stages backups before mutation. It holds one
exclusive project lock, verifies writes, and restores backups after an error.
There is no cross-file filesystem transaction. A process termination can leave
a recovery journal and `.apipak-lock` directory.

First stop other writers and confirm the interrupted R process has exited.
Inspect the recovery plan without writing:

```{r recovery-plan, eval=FALSE}
root <- normalizePath('/path/to/client', winslash = '/', mustWork = TRUE)
apipak::recover_client(root)
```

When no live writer remains, remove only an empty stale lock, if present:

```{r stale-lock, eval=FALSE}
lock <- file.path(root, '.apipak-lock')
if (dir.exists(lock)) {
  stopifnot(
    !fs::is_link(lock),
    length(list.files(lock, all.files = TRUE, no.. = TRUE)) == 0L
  )
  unlink(lock, recursive = TRUE)
}
```

The normalized `root` and literal `.apipak-lock` keep the target explicit. Do
not remove a nonempty directory or use a broad wildcard cleanup. Restore the
reviewed journal, then replan:

```{r recovery-apply, eval=FALSE}
apipak::recover_client(root, 'apply')
apipak::generate_client(root, config = 'apipak.yml', mode = 'plan')
```

Recovery recognizes current and legacy wrapmaint journals, validates destinations
and backups, and retains the journal on failure. Never delete journals/backups
to bypass recovery. The lock is not stolen automatically: directory age does not
distinguish a dead process from a slow live writer.

## Legacy R-list and compatibility interfaces

Existing R-list callers can continue to use the same generator:

```{r legacy, eval=FALSE}
spec <- list(
  files = '/path/to/schema.json', helper = 'request_helper',
  policy_version = 'reviewed-1'
)
apipak::generate_client('/path/to/client', spec, 'plan')
```

Supply exactly one of `spec` and `config`. YAML is the recommended entry point
for new integrations because it keeps service selection and policy explicit.
`render_operation()` and `apply_files()` are lower-level building blocks; most
projects should use `generate_client()` to retain reconciliation and validation.

`bind_tools(group, envir)` installs extracted compatibility functions into an
explicit environment containing client policy. It is intended for existing
endpoint-table integrations. It does not configure a new client automatically.
The client supplies selection, method allowlists, rendering, fixtures, and
reporting policy. Legacy diff counts and helper-call tests are regression
evidence, not a general compatibility proof.

The `readiness` group retains the versioned audit helpers for clients already
using that workflow. Bind into an environment with `asNamespace('apipak')` as
parent and set `audit_policy` through its `read_audit_policy()` reader. Credential
names, issue metadata, badge paths, and test-tier descriptions belong to the
client's audit policy. Prefer the client's existing thin audit command.

## Reporting a reproducible problem

Include the toolkit version, operating system, generation mode, sanitized error,
and the smallest local schema/policy that reproduces it. State which checks ran
and whether any request was sent. Remove keys, tokens, private URLs, and private
response content. Report at the
[apipak issue tracker](https://github.com/seanthimons/apipak/issues).
